R for Clinical Study Reports and Submission

### 2022-08-17 slides/fda-workshop-slides.Rmd

---
title: "R for Clinical Study Reports and Submission"
subtitle: "ASA Princeton-Trenton Chapter"
author: "Yilong Zhang, Nan Xiao"
date: "December 2021"
output:
  mkdocs::mk_ioslides_presentation:
    mk_logo: "invent"
    mk_col_title: "white"
    mk_col_section: "teal"
    mk_class: "public"
    widescreen: true
---

@@

- Delivering TLFs in CSR (Yilong Zhang)
  - Session 1 (45 min)
  - Break and/or Exercise (20 min)
  - Session 2 (45 min)
  - Break (10 min)
- Clinical trial project (Nan Xiao)
  - Session 3 (30 min)
  - Break (10 min)
- eCTD submission package (Nan Xiao)
  - Session 4 (40 min)
  - Q&A (10 min)

@@

- All opinions expressed are those of the presenter and not
  Merck Sharp & Dohme Corp., a subsidiary of Merck & Co., Inc., Kenilworth, NJ, USA.

@@

## Training objective

- Learning how to create tables for clinical study reports.
- Learning how to organize a clinical development project.
- Learning how to prepare eCTD submission package to FDA.

@@

- Team members from Merck & Co.
  - [We are hiring!](https://jobs.merck.com/bards)
  - Contact us if interested.

- [Contributors](https://r4csr.org/) of the training materials.
  - Please consider submitting issues or PR in the repo.

@@

- [ASA Princeton-Trenton Chapter](https://community.amstat.org/princetontrenton/home)
  - Please consider to be a member of our chapter.

<!--

- R / Pharma organizers
  + It is a fun and productive annual gathering.
  + Please consider sharing stories and use cases to expand the community.

-->

@@

- During the workshop, we will use `esubdemo` project.
  - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
  - Project link will be shared in chat
  - Post questions in group chat.

@@

## ICH E3 guidance

The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) provide guidance to assist sponsors in the development of a CSR.

@@

- Public available CDISC pilot [study data located at CDISC Bitbucket repository](https://bitbucket.cdisc.org/projects/CED/repos/sdtm-adam-pilot-project/browse/updated-pilot-submission-package/900172/m5/datasets/cdiscpilot01/analysis/adam/datasets).

@@

- Source data: <https://github.com/elong0527/r4csr/tree/master/data-adam>

@@

# r2rtf introduction

@@

## Minimal example

@@

## Package overview

@@

## Simple example: adverse events

`r2rtf` only focus on **table format**.
Data manipulation and analysis should be handled by other R packages. (e.g., `tidyverse`)

- Let's refer to examples in the [Section 1.4.2](https://r4csr.org/tlf-overview.html#r2rtf)

## Function summary

@@

## Function illustration

@@

# Break and/or exercise (20 min)

# CSR examples

## Disposition table

@@

## Analysis population

@@

## Baseline characteristics

@@

## Efficacy table

@@

## AE Summary table

@@

## Specific AE table

@@

@@

# Clinical trial project

@@

## Folder structure

@@

## Project management

@@

# eCTD submission

<https://r4csr.org/submission-overview.html>

## Submission package

@@

## Running environment

@@

## pkglite: compact package representations

@@

## Pipe-friendly workflow

@@

## File specifications + file collections

@@

>>>>

---
title: "R for Clinical Study Reports and Submission"
subtitle: "The ASA Biopharmaceutical Section Regulatory-Industry Statistics Workshop"
author: "Yilong Zhang, Nan Xiao, Keaven Anderson"
date: "September 2022"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
bibliography: reference.bib
---

@@

- Delivering TLFs in CSR 
  - Session 1 (45 min)
  - Break and/or Exercise (20 min)
  - Session 2 (45 min)
  - Break (10 min)
- Clinical trial project 
  - Session 3 (30 min)
  - Break (10 min)
- eCTD submission package 
  - Session 4 (40 min)
  - Q&A (10 min)

@@

- All opinions expressed are those of the presenter.

@@

## Training Objective

- Learning how to create TLFs for clinical study reports.
- Learning how to organize a clinical development project.
- Learning how to prepare eCTD submission package to FDA.

@@

- [Contributors](https://r4csr.org/preface.html#authors-and-contributors) of the training materials.
  - Please consider submitting issues or PR in the [github repo](https://github.com/elong0527/r4csr).

@@

@@

- During the workshop, we will use `esubdemo` project.
  - `esubdemo` is shared in RStudio Cloud (<https://rstudio.cloud/>)
  - Project link will be shared at the beginning of the course.

As an organization, we need to ensure compliance and reduce the risk of using R and R packages in regulatory deliverables.

- R is widely used in clinical trial study design.
  - Following @lachin1986evaluation
  - [`gsDesign`](https://github.com/keaven/gsDesign): an R package for group sequential design under proportional hazards.
  - [`simtrial`](https://github.com/Merck/simtrial),
  [`gsDesign2`](https://github.com/Merck/gsDesign2), and
  [`gsdmvn`](https://github.com/Merck/gsdmvn): experimental R packages for group sequential design under non-proportional hazards
  - Bookdown: <https://keaven.github.io/gsd-deming/>

- R is flexible for handling novel missing data approaches following the estimand framework in ICH E9 (R1)
  - Recurrent event data: @gao2017control @diao2020efficient
  - Robustness: @liu2021multiply

- R is used in Bayesian statistics.
  - [stan](https://mc-stan.org/users/interfaces/rstan) and network meta-analysis for drug reimbursement analysis

- R is widely used in visualization
  - [SafetyGraphics](https://github.com/SafetyGraphics/safetyGraphics)
  - [forestly](https://elong0527.github.io/forestly/articles/example.html#construct-interactive-forest-plot-1): [Interactive forest plot for DMC safety monitoring in clinical trials](https://rinpharma.com/publication/rinpharma_194/)

@@

## ICH E3 Guidance

The [ICH E3: structure and content of clinical study reports](https://database.ich.org/sites/default/files/E3_Guideline.pdf) 
provides guidance to assist sponsors in the development of a CSR.

@@

- Public available [CDISC pilot study data](https://github.com/elong0527/r4csr/tree/master/data-adam).

@@

@@

# r2rtf Introduction

@@

## Minimal Example

@@

## Package Overview

@@

## Function Summary

@@

## Function Illustration

@@

# Break and/or Exercise (20 min)

# CSR Examples

## Disposition Table

@@

## Analysis Population

@@

## Baseline Characteristics

@@

## Efficacy Table

@@

## AE Summary Table

@@

## Specific AE Table

@@

## Assemble TLFs

<https://r4csr.org/assemble.html>

@@

# Clinical Trial Project

@@

## Folder Structure

@@

## Project Management

@@

# eCTD Submission

## Clarification from FDA

- [FDA: Statistical Software Clarifying Statement](https://www.fda.gov/files/about%20fda/published/Statistical-Software-Clarifying-Statement-PDF.pdf)

> "FDA does not require use of any specific software for statistical analyses, and statistical software
is not explicitly discussed in Title 21 of the Code of Federal Regulations [e.g., in 21CFR part
11]. However, the software package(s) used for statistical analyses should be fully documented
in the submission, including version and build identification."

## R Consortium Pilot Submission Working group

- The R submission working group is a cross-industry pharma working group focusing on improving practices of R-based clinical trial regulatory submission.
  - [members and affiliations](https://rconsortium.github.io/submissions-wg/)
- Mission: Easier R-based clinical trial regulatory submissions
- [pilot 1 - common analyses](https://rconsortium.github.io/submissions-wg/pilot-overall.html#pilot-1---common-analyses)
  - 2021.11   Initial submission to FDA. 
  - 2022.12   Received FDA response 
  - 2022.02   Revised submission to FDA
  - 2022.03   Received final FDA response.

## Challenges and Assumptions

### Challenges

- How to submit internally developed (proprietary) R packages?
- How to follow [ICH and/or FDA guidances](https://www.fda.gov/drugs/electronic-regulatory-submission-and-review/electronic-common-technical-document-ectd)
in preparing eCTD package?
- How to enhance reproducibility from FDA reviewer's perspective?

### Assumptions

- Focus on analysis and reporting given available ADaM datasets.

## Deliverables

- Open source pilot eCTD submission package to FDA
  - Proprietary R package
  - R scripts for analysis
  - Analysis data reviewer guide (ADRG)
  - etc.
- Development: <https://github.com/RConsortium/submissions-pilot1>
- eCTD package: <https://github.com/RConsortium/submissions-pilot1-to-fda>
- [R consortium announcement](https://www.r-consortium.org/blog/2021/12/08/successful-r-based-test-package-submitted-to-fda)

## FDA Response

### Initial Submission

- "Using R version 4.1.1, FDA was able to run the submitted code and confirm the applicant's tables and the submitted figure in report-tlf pdf file."
- "Using FDA developed code, a statistical analyst was able to independently generate tables using the submitted data."
- [Initial submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission.pdf)

### Revised Submission

- FDA agrees that the initial phase of the R Pilot submission has been completed.
- [Revised  submission FDA response](https://github.com/RConsortium/submissions-wg/blob/main/Documents/Summary_R_Pilot_Submission2.pdf)

## Future Work

- Pilot R Shiny app submission to FDA.
- Seek pilot submission opportunities to other regulator agencies. 
- Pilot submission with advanced analysis methods (e.g., study design, missing data, Bayesian, etc).

## Submission Package

@@

## Running Environment

@@

## pkglite: Compact Package Representations

@@

## Pipe-friendly Workflow

@@

## File Specifications + File Collections

@@

## References
